---
title: "Data Visualisation with ggplot2 - Project"
output: html_notebook
---

# AfterWork Data Science: Data Visualisation with Python Project

## 1. Defining the Question

### a) Specifying the Data Analysis Question
Specify the reasearch question that you'll be answering. i.e. Provide strategy recommendations that will lead to revenue growth.

### b) Defining the Metric for Success
The solutions to the following questions will help us answer our research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?

### c) Understanding the context
Provide some background information....

### d) Recording the Experimental Design
Describe the steps/approach that you will use to answer the given question.

### e) Data Relevance
How relevant was the provided data?


## 2. Reading the Data

```{r}
# Load the data below
# --- 
# Dataset url = https://bit.ly/2WvQbhJ
 
# --- 
# YOUR CODE GOES BELOW
Bookings_df<- read.csv("/cloud/project/hotel_bookings.csv")
Bookings_df
```
```{r}
library(tidyverse)
```

```{r}
# Checking the first 5 rows of data
# ---
head(Bookings_df, 5)
```

```{r}
# Checking the last 5 rows of data
# ---
tail(Bookings_df, 5)

```

```{r}
# Sample 10 rows of data
# ---
sample(Bookings_df, 10)
```

```{r}
# Checking number of rows and columns
# ---
dim(Bookings_df)
#  
```

```{r}
# Checking datatypes
# ---
library(dplyr)
glimpse(Bookings_df)
# 
```
The data set has 118,201 rows and 32 columns
Some columns have null values 


## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.


## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# ---
str(Bookings_df)


```
```{r}
colSums(is.na(Bookings_df))
```

We observe the following from our dataset:

Observation 1 
Children variable has 4 missing values

Observation 2
days_in_waiting_list,adr &  total_of_special_requests each has 1 missing value
```{r}
# Checking how many duplicate rows are there in the data
# ---
Bookings_df[duplicated(Bookings_df), ]

```

We choose to keep the duplicates because we don't have a unique identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
colSums(is.na(Bookings_df))
# 
```

We observe the following from our dataset:

Observation 1
Children has 4 missing values

```{r}
# Checking if any of the rows are all null
# ---
is.na(Bookings_df)
# 
```

We observe the following from our dataset:

Observation 1
There are no rows with all null


```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap

install.packages("reshape")
library(reshape)
```


```{r}
# Select the desired columns
Bookings_df <- Bookings_df %>%
na.omit() %>%
  select()

```


```{r}
# Create a correlation matrix
corr_matrix <- cor(Bookings_df, method="s")
head(corr_matrix)

```





```{r}
# Create a table with long form
corr_df <- melt(corr_matrix)
corr_df

```

```{r}
install.packages("ggplot2")
library(ggplot2)
```



```{r}
# Plot the heatmap
ggplot(corr_df, aes(X1, X2, fill = value)) + 
  geom_tile(color = "black") + 
  geom_text(
    aes(label = round(value, 2)), 
    color = "white"
  ) +
  coord_fixed() + 
  labs(
    fill="Pearson Correlation"
  ) +
  scale_fill_gradient2(
    low = "blue", 
    high = "red",
    mid = "white", 
    limit = c(-1,1)
  ) + 
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```



We observe the following from our dataset:
No correlation 


```{r}
glimpse(Bookings_df)
```


```{r}
install.packages("dplyr")
library(dplyr)
```




```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions

library(dplyr)
Bookings_df1 <- Bookings_df
select (Bookings_df,-c(company))
```


From the data variable description we see that the Distribution Channel categoricy that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO
# ---
Bookings_df$distribution_channel[is.na(Bookings_df$distribution_channel)] <- "TO"
Bookings_df

```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
adults <-Bookings_df1$adults
babies <-Bookings_df1$babies
child <-Bookings_df1$children
Bookings_df2 <- subset(Bookings_df, adults!="0" & babies!="0" & child!="0")
head(Bookings_df2)
#Bookings_df2 <- Bookings_df1[-c(adults == 0), ]
```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
Bookings_df2$children[is.na(Bookings_df2$children)] <- mean(dt$age, na.rm = TRUE)
Bookings_df2
```

```{r}
# Checking for missing values in the dataframe
# ---
is.na(Bookings_df2)
#
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW
lapply(Bookings_df2$children,as.numeric)
lapply(Bookings_df2$company,as.numeric)
lapply(Bookings_df2$agent,as.numeric)
#  Bookings_df <- Bookings_df %>% 
#     mutate_at(c(11, 24, 25), as.integer())
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# Visualisation: Barplot
# Create the summary table
cancelled_df <- Bookings_df2 %>%
  group_by(hotel) %>%
summarise(is_canceled = sum(is_canceled))
```

```{r}
# Plotting Barplot chart to visualize the data
ggplot(cancelled_df, aes(x = hotel, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Cancelled Bookings per Hotel", 
    x = "hotel", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +
  geom_text(aes(label = is_canceled, face = "bold"), vjust = 1.5, colour = "black")
```
City hotel had 13 cancelled bookings while resort hotel had 20 cancelled bookings

```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# Create the summary table
ratio_df <- Bookings_df %>%
  group_by(hotel) %>%
  summarise(is_canceled = sum(is_canceled==0))



# Barplot of booking ratio between resort hotel and city hotel
ggplot(ratio_df, aes(x = hotel, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Hotel", 
    x = "hotel", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```

City hotel had a ratio of 60.9% and resort hotel had a ratio of 39.1%

```{r}
# 3. What was the percentage of booking for each year?
year_df <- Bookings_df %>%
  group_by(arrival_date_year) %>%
  summarise(is_canceled = sum(is_canceled==0))

# Plotting Barplot chart to visualize the data
# Plotting Barplot chart to visualize the data
ggplot(year_df, aes(x = arrival_date_year, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Year", 
    x = "year", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 0, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```
2016 had the highest bookings at 49.2% followed by 2017 at 32.1% and finally 2015 at 18.7% 


```{r}
# 4. Which were the most busiest months for hotels?
month_df <- Bookings_df %>%
  group_by(arrival_date_month) %>%
  summarise(is_canceled = sum(is_canceled==0))


# Plotting Barplot chart to visualize the data
ggplot(month_df, aes(x = arrival_date_month, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Month", 
    x = "month", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")

```
Most bookings are in the months of July and August

```{r}
# 5. From which top 3 countries did most guests come from?
countries_df <- Bookings_df %>%
  group_by(country) %>%
  summarise(is_canceled = sum(is_canceled==0))
countries_df
arrange(countries_df,desc(is_canceled))

#Filtering to get the top three countries
top3_countries_df <- filter(countries_df, country == "PRT" | country == "GBR" | country == "FRA")
top3_countries_df

# Plotting Barplot chart to visualize the data
ggplot(top3_countries_df, aes(x = country, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Month", 
    x = "country", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```
PRT had the highest number of customers booking, followed by GBR and FRA in the third position



```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel

stays_df <- Bookings_df %>%
  mutate(total_stays = stays_in_weekend_nights + stays_in_week_nights) %>%
  group_by(total_stays) %>%
  summarise(is_canceled = sum(is_canceled==0))
  arrange(stays_df, desc(is_canceled))
stays_df

# Plotting Barplot chart to visualize the data
ggplot(stays_df, aes(x = total_stays)) +
  geom_histogram(
    bins = 10,fill = "#0099f9"
  ) +
  labs(
    title = "Sum of Total stays in nights", 
    x = "total_stays", 
    y = "Frequency"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )+
  scale_x_continuous(breaks = seq(0, 50,by=5))
```
Most people stay between 0 to 20 days


```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
accommodation_df <- Bookings_df %>%
  mutate(hotel_accommodation =
           case_when(
             Bookings_df$adults >= 0 & (Bookings_df$children >= 1 | Bookings_df$babies >= 1) ~ "Family",
             Bookings_df$adults == 2 & Bookings_df$children == 0 & Bookings_df$babies == 0 ~ "Couple",
             Bookings_df$adults == 1 & Bookings_df$children == 0 & Bookings_df$babies == 0 ~ "Single",
            Bookings_df$adults > 2 & (Bookings_df$children == 0 | Bookings_df$babies == 0) ~ "Family")
  )%>%
  filter(is_canceled == 0)
  
accommodation_df1 <- accommodation_df %>%
  group_by(hotel_accommodation) %>%
  count(hotel_accommodation)

#Plotting the bar chart
ggplot(accommodation_df1, aes(x = hotel_accommodation, y = n, fill = accommodation_df)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Sum of Hotel Accommodataion by accomodation type",
    x = "Hotel Accommodation", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
Couples make up the largest bookings, followed by singles and Family


### 5.b) Recommendations
From the above analysis, below are our recommendations:

More family products should be tailored so as to encourage more bookings from that segment
Offers and discounts around long stays should be considered as longer stays increase revenue
Capacity building should be done for the months of July and August to cater for the many bookings


## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.

```{r}
# When should hotels provide special offers?
month_df <- Bookings_df %>%
  group_by(arrival_date_month) %>%
  summarise(is_canceled = sum(is_canceled==0))
# Plotting Barplot chart to visualize the data
ggplot(month_df, aes(x = arrival_date_month, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Bookings per Month", 
    x = "month", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  geom_text(aes(label = paste0(round(is_canceled / sum(is_canceled) * 100, 1), "%")), vjust = 1.5, colour = "black")
```
Give offers in rates In the months of January, December and November to try and boost bookings


Our observations:
-

How does this observation tie to our solution?

## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data?YES we did
b). Do we need other data to answer our question?No, the data was sufficient
c). Did we have the right question?Yes we did 


